Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-06 manuscript/paper.tex

Visualisation
is an essential component of data analysis pipelines. In the past few decades,
genome browsers have become synonymous with sequence visualisation, by
providing user-friendly, interactive interfaces to navigate genomic sequences
and annotations. From early standalone tools to manage \textit{C. elegans} data
\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
with the changing landscape of genomics. Modern genome browsers can load
datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
The data security and scalability offered by this architecture ensures that
genome browsers continue to be relevant in the biobank era.

@@

To address this gap, we have developed \texttt{tsbrowse}, an open-source app
intended as a genome browser for ARGs. Its  focus is on visualising the
information structure in ARGs, thereby improving  inference and analysis at
biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
\citep{wong2024general}. This is the native format of \texttt{tskit} \citep{tskit}, a
widely-used library for interchanging ARGs inferred using a variety of methods
\citep{speidel2019method,zhang2023biobank,Haller2019,Baumdicker2021,gunnarsson2024scalable,
deng2024robust}.
The interface is run on a web browser. It works on biobank-scale datasets
without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
for performing quality control and exploratory analyses on ARGs inferred from
biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

@@

\subsection{Data model} \label{subsec:Data_Model} The foundation of
\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
based on this generalised ARG representation \citep{wong2024general} it can be applied
to ARGs from several widely-used inference methods including \texttt{Relate}
\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} 
and \texttt{SINGER} \citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
collection of tables, each of which store information about a specific ARG
property. For example, the Node table describes sampled or ancestral sequences
and records their time of birth, and the individual and population they belong
to. The Edge table describes the relationship between pairs of nodes, and
records the node IDs and the genomic locations over which the edge spans. The
Mutation table records mutation events in the history of the sample, including
the change of allelic state, time of event and node in which the event
occurred. The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, while the Mutation table describes
sequence variation contained in the samples \citep{Ralph2020}. This compact
representation enables the efficient storage and querying of biobank-scale
ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation (see
\hyperref[subsec:User_Interface]{User Interface}).

@@

 Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2}
highlights a use-case in visually characterising ARGs. The input ARG is a
selective sweep dataset simulated using \texttt{msprime}
\citep{Baumdicker2021}. The distinct pattern of edges observed in the middle of
the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst
in recent coalescent events, showing that tsbrowse captures this characteristic
of a selective sweep. Supplementary Figure \ref{fig:Supplementary_Figure_1} is
another example of characterising ARGs with tsbrowse. The view of edges from
four different inferences of the same dataset (inferred with \texttt{tsinfer} +
\texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle}
(Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences
between inference methods.

@@

Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
Mutations page for SARS-CoV-2 ARGs \citep{Zhan2023.06.08.544212} demonstrates
tsbrowse's ability to handle diverse data sources. This plot represents a total
of \textcolor{red}{X million} mutations from an inference of
\textcolor{red}{1.27 million} sequences. After executing the
\texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
large-scale datasets in a performant way. Panel B demonstrates a useful
interactive feature; additional information about a mutation event (for
example, the mutational frequency in predefined populations, or the number of
sample nodes carrying the mutation) is displayed on mouse-over of individual
data points.

@@

In current practice, simulated ARGs are used as ground-truth to validate
inferences from empirical data
\citep{kelleher2019inferring,zhang2023biobank,speidel2019method,
gunnarsson2024scalable}.
Sophisticated techniques
\citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
under complex, realistic demographic models \cite{Anderson-Trocme2023} now
exist. However, models of simulation do not capture \textit{all} the nuances of
population datasets. This results in a clear gap between ARG inference and
application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
providing the means to qualitatively assess inferences. With notable features
of browsability along the genome, scalability and applicability to diverse data
sources, inference and simulation methods, \texttt{tsbrowse} has the potential
to be integrated into biobank-scale data pipelines to improve ARG inference and
analysis.

@@

The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with
\texttt{sc2ts}~\citep{Zhan2023.06.08.544212} using the Viridian
dataset~\citep{hunt2024addressing}. It consists of 516,443 samples, 612,209
nodes, 612,655 edges and 988,303 mutations.

>>>>

Visualisation
is an essential component of data analysis pipelines. In the past few decades,
genome browsers have become synonymous with sequence visualisation, by
providing user-friendly, interactive interfaces to navigate genomic sequences
and annotations. From early standalone tools to manage \textit{C. elegans} data
\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
with the changing landscape of genomics. Modern genome browsers can load
datasets from a URL or local file system \citep{Lee2013,Robinson2023}.
The data security and scalability offered by this architecture ensures that
genome browsers continue to be relevant in the biobank era.

@@

To address this gap, we have developed \texttt{tsbrowse}, an open-source app
intended as a genome browser for ARGs. Its  focus is on visualising the
information structure in ARGs, thereby improving  inference and analysis at
biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
\citep{wong2024general}. 
This the \texttt{tskit} library~\citep{ralph2020efficiently}
and supported by 
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,tsambos2023link,tagami2024tstrait}
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
and downstream processing methods~\citep{link2023tree,nowbandegani2023extremely}

% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.

% widely-used library for interchanging ARGs inferred using a variety of methods
% \citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.

The interface is run on a web browser. It works on biobank-scale datasets
without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
for performing quality control and exploratory analyses on ARGs inferred from
biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

@@

\subsection{Data model} \label{subsec:Data_Model} The foundation of
\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
based on this generalised ARG representation \citep{wong2024general} it can be applied
to ARGs from several widely-used inference methods including \texttt{Relate}
\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} and \texttt{SINGER}
\citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
collection of tables, each of which store information about a specific ARG
property. For example, the Node table describes sampled or ancestral sequences
and records their time of birth, and the individual and population they belong
to. The Edge table describes the relationship between pairs of nodes, and
records the node IDs and the genomic locations over which the edge spans. The
Mutation table records mutation events in the history of the sample, including
the change of allelic state, time of event and node in which the event
occurred. The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, while the Mutation table describes
sequence variation contained in the samples. This compact
representation enables the efficient storage and querying of biobank-scale
ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation (see
\hyperref[subsec:User_Interface]{User Interface}).

@@

\subsection{Applications}

 Tsbrowse can be applied to versatile scenarios. Figure \ref{fig:Figure_2}
highlights a use-case in visually characterising ARGs. The input ARG is a
selective sweep dataset simulated using \texttt{msprime}
\citep{baumdicker2022efficient}. The distinct pattern of edges observed in the middle of
the sequence between \textit{y=0} to \textit{y=2.5} is attributable to a burst
in recent coalescent events, showing that tsbrowse captures this characteristic
of a selective sweep. Supplementary Figure \ref{fig:Supplementary_Figure_1} is
another example of characterising ARGs with tsbrowse. The view of edges from
four different inferences of the same dataset (inferred with \texttt{tsinfer} +
\texttt{tsdate} (Panel A), \texttt{Relate} (Panel B), \texttt{ARG-Needle}
(Panel C) and \texttt{SINGER} (Panel D)), reflects qualitative differences
between inference methods.

@@

Supplementary Fig. \ref{fig:Supplementary_Figure_4}, a screenshot of the
Mutations page for SARS-CoV-2 ARGs \citep{zhan2023towards} demonstrates
tsbrowse's ability to handle diverse data sources. This plot represents a total
of \textcolor{red}{X million} mutations from an inference of
\textcolor{red}{1.27 million} sequences. After executing the
\texttt{preprocess} step, the page loads in \textcolor{red}{X} seconds on a
system with \textcolor{red}{Y GB RAM}, showing that it is possible to visualise
large-scale datasets in a performant way. Panel B demonstrates a useful
interactive feature; additional information about a mutation event (for
example, the mutational frequency in predefined populations, or the number of
sample nodes carrying the mutation) is displayed on mouse-over of individual
data points.

@@

% In current practice, simulated ARGs are used as ground-truth to validate
% inferences from empirical data
% \citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
% Sophisticated techniques
% \citep{haller2019tree,baumdicker2022efficient,adrion2020community,lauterbur2023expanding} to simulate ARGs
% under complex, realistic demographic models \cite{Anderson-Trocme2023} now
% exist. However, models of simulation do not capture \textit{all} the nuances of
% population datasets. This results in a clear gap between ARG inference and
% application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
% providing the means to qualitatively assess inferences. With notable features
% of browsability along the genome, scalability and applicability to diverse data
% sources, inference and simulation methods, \texttt{tsbrowse} has the potential
% to be integrated into biobank-scale data pipelines to improve ARG inference and
% analysis.

@@

The ARG shown in \ref{fig:Supplementary_Figure_4} was inferred with
\texttt{sc2ts}~\citep{zhang2023biobank} using the Viridian
dataset~\citep{hunt2024addressing}. It consists of 516,443 samples, 612,209
nodes, 612,655 edges and 988,303 mutations.
